---
title: "Exploratory Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install, include=FALSE}
# Install required packages
packages = c('dplyr', 'ggplot2', 'tidyverse', 'outliers', 'here')

# install packages if needed
for (package in packages) {
  if (!package %in% installed.packages()) {
    install.packages(package)
  }
}
```

```{r load-libraries}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyverse)
library(outliers)
library(here)
```

# Load the Data

```{r}
glassdoor_data <- read.csv(here("Data", "Glassdoor Gender Pay Gap.csv"), header = TRUE)
```

## View the First Few Rows

```{r}
head(glassdoor_data)
```

## Data Summary

```{r}
# view the data and dtypes
str(glassdoor_data)
summary(glassdoor_data)
```

## Missing Data

```{r}
any(is.na(glassdoor_data))
```

## Record Counts by Gender

```{r}
glassdoor_data %>%
  group_by(Gender) %>%
  summarise(count = n()) %>%
  mutate(perc_total = count / sum(count) * 100)
```

## Record Counts by Education

```{r}
glassdoor_data %>%
  group_by(Education) %>%
  summarise(count = n()) %>%
  mutate(perc_total = count / sum(count) * 100) 
```

## Record Counts by Department

```{r}
glassdoor_data %>%
  group_by(Dept) %>%
  summarise(count = n()) %>%
  mutate(perc_total = count / sum(count) * 100) 
```

## Record Counts by Seniority

```{r}
glassdoor_data %>%
  group_by(Seniority) %>%
  summarise(count = n()) %>%
  mutate(perc_total = count / sum(count) * 100) 
```

Record count/sample appears to be balanced when looking at the data based on: Edu, Dept, & Seniority. Gender is somewhat balanced with a slight majority of male records.

## Find Outliers

### AGE

```{r}
grubbs.test(glassdoor_data$Age)
boxplot(glassdoor_data$Age)
```

p = 1  
alternative hypothesis: highest value 65 is an outlier  
no points beyond the "whiskers" of the box plot  

```{r}
grubbs.test(glassdoor_data$PerfEval)
boxplot(glassdoor_data$PerfEval)
```

p = 1  
alternative hypothesis: lowest value 1 is an outlier  
no points beyond the "whiskers" of the box plot  

```{r}
grubbs.test(glassdoor_data$Seniority)
boxplot(glassdoor_data$Seniority)
```

p = 1  
alternative hypothesis: highest value 5 is an outlier  
no points beyond the "whiskers" of the box plot  

### BASEPAY

```{r}
grubbs.test(glassdoor_data$BasePay)
boxplot(glassdoor_data$BasePay)
outliers_basepay <- boxplot(glassdoor_data$BasePay)$out

# Print the outlier values
cat("Outlier values: ", outliers_basepay, "\n")
```

p = 0.3711  
alternative hypothesis: highest value 179726 is an outlier  
THERE ARE points beyond the "whiskers" of the box plot  
to find the values:  
Outlier values:  176789 165229 179726   

### BONUS

```{r}
grubbs.test(glassdoor_data$Bonus)
boxplot(glassdoor_data$Bonus)
```

p = 1  
alternative hypothesis: highest value 11293 is an outlier  
no points beyond the "whiskers" of the box plot  

## Histogram for Age

```{r}
ggplot(glassdoor_data, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "red") +
  labs(title = "Age Histogram", x = "Age")
```

## Box Plots for Base Pay

```{r}
glassdoor_data %>%
  gather(variable, value, BasePay) %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  labs(title = "BasePay Box Plot", x = "Base Pay Variable", 
       y = "Amount Earned")
```

## Box Plots for Bonus

```{r}
glassdoor_data %>%
  gather(variable, value, Bonus) %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  labs(title = "Bonus Box Plot", x = "Bonus Variable", 
       y = "Amount Earned")
```

## Scatter Plot: Age vs. Base Pay

```{r}
ggplot(glassdoor_data, aes(x = Age, y = BasePay)) +
  geom_point(color = "red") +
  labs(title = "Age vs. BasePay", x = "Age", y = "Base Pay")
```
