---
title: "Linear Regression Analysis"
output: html_document
date: "2023 Fall Semester"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(ggplot2)
library(dplyr)
library(here)
library(car)
library(MASS)
```

## Load Data without Outliers
```{r data preparation}
data_no_outlier <- read.csv(here("Data", "Glassdoor Gender Pay Gap w_o outliers.csv"))
summary(data_no_outlier)

# Combine BasePay & Bonus to Salary variable
data_no_outlier$Salary <- data_no_outlier$BasePay + data_no_outlier$Bonus
data_no_outlier <- subset(data_no_outlier, select = -c(BasePay, Bonus))
```

## Separate data for training & testing

We will use 75% of the data for training and 25% for testing.

```{r separate data}
set.seed(123)
training_indices <- sample(nrow(data_no_outlier), 0.75 * nrow(data_no_outlier))

training_data <- data_no_outlier[training_indices, ]
testing_data <- data_no_outlier[-training_indices, ]
```

## Check for Multicollinearity
```{r}
cor(training_data[, c("Age", "PerfEval", "Seniority", "Salary")])
```

Our results indicate a good positive correlation between Age & Salary, suggesting that older employees tend to have higher salaries.
However, Seniority & PerfEval have a lower correlation with Salary, suggesting that these variables may not be good predictors of Salary.

```{r}
# Model all predictors against Salary
model <- lm(Salary ~ ., data = training_data)
summary(model)
```
Gender does not appear to be a statistically significant predictor.

## Check VIF
```{r}
vif(model)
```
Despite some hgih GVIF values, this may not be an issue as GVIF^(1/(2*Df)) < 2.5 for all predictors.

## Check Residuals
```{r}
predictor <- as.factor(training_data$Gender)

plot(predictor, resid(model),
    main = "Residual Plot",
    xlab = "Gender",
    ylab = "Residuals",
    pch = 16, col = "blue")
```

There does not appear to be any noticable pattern in the residuals.

## Isolate predictors of significance
```{r}
start_model <- lm(Salary ~ ., data = training_data)

while(TRUE) {
    model <- stepAIC(start_model, direction = "backward", trace = TRUE)

    if (AIC(model) < AIC(start_model))
        start_model <- model

    break
}
summary(model)
```
As noted earlier, because Gender does not appear to be statistically significant, it was removed from the model using stepwise model selection.

## Measure Performance
Since we are working to identify potential Gender Pay Gap, we will recreate the model with all predictors.
```{r}
model <- lm(Salary ~ ., data = training_data)
summary(model)

predicted_salary <- predict(model, testing_data)
plot(testing_data$Salary, predicted_salary,
    main = "Actual vs. Predicted Salary",
    xlab = "Actual Salary",
    ylab = "Predicted Salary",
    pch = 16, col = "blue")
abline(a = 0, b = 1, col = "red")

# Calculate RMSE
sqrt(mean((testing_data$Salary - predicted_salary)^2))

```

RMSE indicates a moderate error rate of ~$10,500.
In conclusion, while the model performs well, Gender does not appear to be a statistically significant predictor of Salary.

# Linear Regression Analysis per Category
## Department

```{r}
dept_model <- lm(Salary ~ Dept * Gender, data = training_data)
summary(dept_model)
```
Per p-values, Gender does not play a significant role in Department Salaries.

## JobTitle

```{r}
job_model <- lm(Salary ~ JobTitle * Gender, data = training_data)
summary(job_model)
```
Marketing Associate (Male) appears to have a statistically significant impact on Salary.

## Education
```{r}
education_model <- lm(Salary ~ Education * Gender, data = training_data)
summary(education_model)
```
Per p-values, Gender does not play a significant role in Education Salaries.

## Age
```{r}
age_model <- lm(Salary ~ Age * Gender, data = training_data)
summary(age_model)
```
Per p-values, Gender does not play a significant role in Age Salaries.

## Seniority
```{r}
seniority_model <- lm(Salary ~ Seniority * Gender, data = training_data)
summary(seniority_model)
```
Per p-values, Gender does not play a significant role in Seniority Salaries.

## PerfEval
```{r}
perfeval_model <- lm(Salary ~ PerfEval * Gender, data = training_data)
summary(perfeval_model)
```
Per p-values, Gender does not play a significant role in PerfEval Salaries.

## Deep-dive Job Title Marketing Associate
```{r}
marketing_model <- lm(Salary ~ Gender + Age, data = subset(training_data, JobTitle == "Marketing Associate"))
summary(marketing_model)
```
There appears to be statistical significance regarding Gender + Age for Marketing Associates, however, the adjusted R2 value is low.
The p-values may be due to our previous findings that Age is a good predictor of Salary.

```{r}
predicted_salary <- predict(marketing_model, subset(testing_data, JobTitle == "Marketing Associate"))
plot(subset(testing_data, JobTitle == "Marketing Associate")$Salary, predicted_salary,
    main = "Actual vs. Predicted Salary",
    xlab = "Actual Salary",
    ylab = "Predicted Salary",
    pch = 16, col = "blue")

abline(a = 0, b = 1, col = "red")

# Calculate RMSE
sqrt(mean((subset(testing_data, JobTitle == "Marketing Associate")$Salary - predicted_salary)^2))
```
RMSE indicates a moderate error rate of ~$20,000.
In conclusion, even after diving deeper into JobTitles with potential Gender Pay Gap in Marketing Associates, the model performs poorly.
