---
title: "Logistic Regression Model Validation"
output: html_document
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. Cross - Validation:
```{r}
library(boot)

# Logistic Model
cv_results <- cv.glm(data, logistic_model, K = 10)  # 10-fold cross-validation
print(cv_results)

# Interaction Model
cv_results_2 <- cv.glm(data, logistic_interaction_model, K = 10)  # 10-fold cross-validation
print(cv_results_2)

```

Two delta values are found for the logistic model: 0.09246339 0.09206775. For the interaction model: 0.09534866 0.09450217. Both are performing better in terms of deviance reduction due to low delta value. 

2. ROC Curve
```{r}
library(pROC)
roc_curve <- roc(data$HighSalary, predicted_probs)
auc_value <- auc(roc_curve)
plot(roc_curve, main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"))
```

An AUC of 0.96 indicates that the model is making accurate predictions and has a high level of sensitivity and specificity.

3. Residual Analysis
```{r}
residuals <- residuals(logistic_model, type = "response")

# Residuals vs. predicted values
plot(predicted_probs, residuals, xlab = "Predicted Probabilities", ylab = "Residuals")

# Residuals vs. predictor variable (replace 'Age' with your predictor variable)
plot(data$Age, residuals, xlab = "Age", ylab = "Residuals")

```

Residuals are distributed symmetrically around zero. Positive residuals and negative residuals are balanced.The residuals generally cluster around zero. 